---
title: "Genome-wide methylation analysis using coMethDMR via parallel computing"
author: "Gabriel J. Odom, Lissette Gomez, Tiago Chedraoui Silva, Lanyu Zhang and Lily
  Wang"
date: "March 2021"
output: 
    BiocStyle::html_document:
    toc_float: true
    toc: true
    df_print: paged
    code_download: false
    toc_depth: 3
editor_options:
  chunk_output_type: inline   
vignette: >
  %\VignetteIndexEntry{"coMethDMR with Parallel Computing"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
In the previous vignette “Introduction to **coMethDMR**”, we discussed components of the **coMethDMR** pipeline and presented example R scripts for running an analysis with **coMethDMR** serially. However, because identifying co-methylated clusters and fitting mixed effects models on large numbers of genomic regions can be computationally expensive, we illustrate implementation of parallel computing for **coMethDMR** via the **BiocParallel** R package in this vignette. 


## Installation 

The latest version can be installed by:
```{r eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("coMethDMR")
```

First, we load these two packages.
```{r packLoad, message=FALSE}
library(coMethDMR)
library(BiocParallel)
```


## Overview
In Section 2, we give a brief re-introduction to the example data. In Section 3, we present example scripts for analyzing a single type (e.g. genic regions) of genomic regions using parallel computing. In Section 4, we present example scripts for analyzing genic and intergenic regions on the Illumina arrays using parallel computing. 



# Example Dataset
For illustration, we use a subset of prefrontal cortex methylation data (GEO GSE59685) described in Lunnon et al. (2014). This example dataset includes beta values for 8552 CpGs on chromosome 22 for a random selection of 20 subjects. We assume quality control and normalization of the methylation dataset have been performed by R packages such as **minfi** or **RnBeads**.
```{r ex_data}
data(betasChr22_df)
betasChr22_df[1:5, 1:5]
```

The corresponding phenotype dataset included variables `stage` (Braak AD stage), `subject.id`, `slide` (batch effect), `sex`, `Sample` and `age.brain` (age of the brain donor). 
```{r ex_pheno_data}
data(pheno_df)
head(pheno_df)
```

Note that only samples with both methylation data and non-missing phenotype data are analyzed by **coMethDMR**. So in this example, the sample with `subject.id = 3` which lacks AD stage information will be excluded from analysis. Please also note the phenotype file needs to have a variable called "Sample" that will be used by **coMethDMR** to link to the methylation dataset. 



# Analyzing One Type of Genomic Region via BiocParallel
As mentioned previously in “Introduction to **coMethDMR**”, there are several steps in the **coMethDMR** pipeline:

1. Obtain CpGs located closely in pre-defined genomic regions,
2. Identify co-methylated regions, and
3. Test co-methylated regions against the outcome variable (AD stage).

Suppose we are interested in analyzing genic regions on the 450k arrays. In the following, `closeByGeneAll_ls` is a list, where each item includes at least 3 CpGs located closely (max distance between 2 CpGs is 200bp by default). 
```{r list_of_cgis}
closeByGeneAll_ls <- readRDS(
  system.file(
    "extdata",
    "450k_Gene_3_200.rds",
    package = 'coMethDMR',
    mustWork = TRUE
  )
)
```

We can inspect the first region in the list via:
```{r}
closeByGeneAll_ls[1]
```

Next, for demonstration, we select regions on chromosome 22. 
```{r}
indx <- grep("chr22:", names(closeByGeneAll_ls))

closeByGene_ls <- closeByGeneAll_ls[indx]
rm(closeByGeneAll_ls)
length(closeByGene_ls)
```
There are 676 genic regions to be tested for chromosome 22. As an example, this vignette will analyze the first 10 regions, which contain the following CpGs:
```{r}
closeByGene_ls[1:10]
```

In order to make these examples as simple as possible, we will remove the CpGs from our data set which are not included in these 10 regions. In practice, you would keep the full data set.
```{r subset_betasChr22}
keepCpGs_char <- unique(unlist(closeByGene_ls[1:10]))
betasChr22small_df <- 
  betasChr22_df[rownames(betasChr22_df) %in% keepCpGs_char, ]
dim(betasChr22small_df)
```


## Compute residuals
This step removes uninteresting technical and biological effects using the `GetResiduals` function, so that the resulting co-methylated clusters are only driven by the biological factors we are interested in. Note that there may be an error with the parallel back-end (`1 parallel job did not deliver a result`); this has occurred rarely for us (and not consistently enough for us to create a reproducible example), but it goes away when we execute the code a second time. We then recommend that you test this with a smaller subset of your data first. If you are able to consistently replicate a parallel error, please submit a bug report: <https://github.com/TransBioInfoLab/coMethDMR/issues>.
```{r, results='hide'}
resid_df <- GetResiduals(
  dnam = betasChr22small_df,
  # converts to Mvalues for fitting linear model
  betaToM = TRUE,  
  pheno_df = pheno_df,
  # Features in pheno_df used as covariates
  covariates_char = c("age.brain", "sex", "slide"),
  nCores_int = 2
)
```


## Finding co-methylated regions

The cluster computing with the **BiocParallel** package involves simply changing the default value of one argument: `nCores_int`. This argument enables you to perform parallel computing when you set the number of cores to an integer value greater than 1. *If you do not know how many cores your machine has, use the `detectCores()` function from the `parallel` package. Note that, if you have many applications open on your computer, you should not use all of the cores available.*

Once we know how many cores we have available, we execute the `CoMethAllRegions()` function using each worker in the cluster, to find co-methylated clusters in the genic regions. As an example, we only analyze the first 10 CpG regions (`CpGs_ls`).
```{r BiocParallel_Gene, message=FALSE}
system.time(
  coMeth_ls <- CoMethAllRegions(
    dnam = resid_df,
    betaToM = FALSE,
    method = "spearman",
    arrayType = "450k",
    CpGs_ls = closeByGene_ls[1:10],
    nCores_int = 2
  )
)
# Windows: NA
# Mac: ~14 seconds for 2 cores
```

The object `coMeth_ls` is a list, with each item containing the list of CpGs within an identified co-methylated region. 

## Testing the co-methylated regions
Next we test these co-methylated regions against continuous phenotype `stage`, adjusting for covariates `age` and `sex`, by  executing the `lmmTestAllRegions()` function.  
```{r singleRegionType_lmmTest, warning=FALSE, results='hide', message = FALSE}
res_df <- lmmTestAllRegions(
  betas = betasChr22small_df,
  region_ls = coMeth_ls,
  pheno_df = pheno_df,
  contPheno_char = "stage",
  covariates_char = c("age.brain", "sex"),
  modelType = "randCoef",
  arrayType = "450k",
  nCores_int = 2
  # outLogFile = "res_lmm_log.txt"
)
```

Model fit messages and diagnostics for each region will be saved to the log file specified with the `outLogFile` argument. For a single region, this will return a one row of model fit statistics similar to the following:

| chrom |    start |      end | nCpGs | Estimate | StdErr |    Stat | pValue |
|-------|----------|----------|-------|----------|--------|---------|--------|
| chr22 | 24823455 | 24823519 |      4|  -0.0702 | 0.0290 | -2.4184 | 0.0155 |

Finally, we can annotate these results:
```{r}
AnnotateResults(res_df)
```



# coMethDMR Analysis Pipeline for 450k Methylation Arrays Datasets via BiocParallel
In this section, we provide example scripts for testing genic and intergenic regions using **coMethDMR** and **BiocParallel** R packages. 

First, we read in clusters of CpGs located closely on the array, in genic and intergenic regions, then combine them into a single list.
```{r}
closeByGene_ls <- readRDS(
  system.file(
    "extdata",
    "450k_Gene_3_200.rds",
    package = 'coMethDMR',
    mustWork = TRUE
  )
)

closeByInterGene_ls <- readRDS(
  system.file(
    "extdata",
    "450k_InterGene_3_200.rds",
    package = 'coMethDMR',
    mustWork = TRUE
  )
)

# put them together in one list
closeBy_ls <- c(closeByInterGene_ls, closeByGene_ls)
length(closeBy_ls)
closeBy_ls[1:3]
```

With this list of regions and their probe IDs, repeat the analysis code in the previous section.



# A Comment on Using EPIC Methylation Arrays Datasets 
The analysis for EPIC methylation arrays would be the same as those for 450k arrays, except by testing genomic regions in files "EPIC_Gene_3_200.RDS" and "EPIC_InterGene_3_200.RDS", instead of "450k_Gene_3_200.RDS" and "450k_InterGene_3_200.RDS". These data sets are available at <https://github.com/TransBioInfoLab/coMethDMR_data/tree/main/data>.



# Additional Comments on Computational Time and Resources
In this vignette, we have analyzed a small subset of a real EWAS dataset (i.e. only chromosome 22 data on 20 subjects). To give users a more realistic estimate of time for analyzing real EWAS datasets, we also measured time used for analyzing the entire Lunnon et al. (2014) dataset with 110 samples on all chromosomes. These computation times measured on a Dell Precision 5810 with 64Gb of RAM, an Intel Xeon E5-2640 CPU at 2.40Ghz, and using up to 20 cores. More specifically, in Section 4, the entire **coMethDMR** workflow for took 103 minutes with 6 cores and used a maximum of 24Gb of RAM (for the `CoMethAllRegions()` function). We’re currently working improving the speed and reducing the size of **coMethDMR**, so please check back soon for updates. 



# Session Information
```{r}
utils::sessionInfo()
```

